---
title: "Lab 4 - In class exercise - Answers"
author: "Luis Navarro"
output:
  html_document:
    df_print: paged
---




1.	Import the fastfood data set explored in class.  Load the packages dplyr, ggplot2, and descriptr 

```{r}
# Clean the environment 
rm(list=ls())

# Set your working directory
setwd("/Users/luisenriquenavarro/Library/CloudStorage/OneDrive-IndianaUniversity/V506/Fall24/Lab4")

# New Packages
if(!require(descriptr)){install.packages("descriptr")}

# Packages Required for the session 
library(pacman)
p_load(dplyr, ggplot2, rmarkdown, 
       descriptr, 
       rio, here)

# Load fast food
fastfood <- rio::import(file = "fastfood.csv", header = TRUE) %>% tibble() 
```


2.	Following the code covered in class, write a function that takes as input a vector of data (say a variable from a data frame) The function should compute the mean, standard deviation, variance, and median of such variable. The output of the function should be a data frame with 4 columns: Mean, Median, SD, and Variance. 


There are two potential answers for this. Using base R and summarize from dplyr. 

```{r}
# Base R
descriptive_stats <- function(data){
  
  # Compute the Descriptive Statistics using base R functions
  mean <- mean(data, na.rm = TRUE)
  median <- median(data, na.rm = TRUE)
  sd <- sd(data, na.rm = TRUE)
  var <- var(data, na.rm = TRUE)
  
  # Save output as data frame (tibble object)
  # note: I capitalize the variable names just to avoid confusion with the objects defined inside the function. 
  # you can assign them the same name as the objects and it works fine 
  table <- tibble(
    Mean = mean,
    Median = median,
    SD = sd,
    VAR = var
    ) 

  # Return Table 
  return(table)
}
```


```{r}
# test the function 
descriptive_stats(
  data = fastfood$calories
  )

```

```{r}
# solution 2: using summarize from dplyr 
desc_table2 <- function(data){
  # the key difference here is that data is not a vector, but a data frame 
  # verify the data used as input is a data frame object (tibble). 
  df <- tibble(variable = data)
  # create the table with desc stats
  table <- df %>% 
    summarize(
      Mean = mean(variable, na.rm = TRUE), 
      Median = median(variable, na.rm = TRUE), 
      SD = sd(variable, na.rm = TRUE), 
      VAR = var(variable, na.rm = TRUE)
    ) 
  
  return(table)
}

```


```{r}
# test the function 
desc_table2(
  data = fastfood$calories
)

```



3.	Use your user written function to obtain the descriptive statistics of the variables: i) calories, ii) cholesterol, iii) protein, and iv) sugar. Use lapply and bind_rows to visualize the output in one data frame. Add a column with the variable names to identify the output of each row. 

```{r}
# using base R
data_list <- list(fastfood$calories, 
                  fastfood$cholesterol, 
                  fastfood$protein, 
                  fastfood$sugar)

# we can use lapply 
data_summaries <- lapply(X = data_list, 
                         FUN = desc_table2)

# merge them using bind_rows
data_summaries_table <- bind_rows(
  data_summaries[[1]] %>% mutate(var = "calories"), 
  data_summaries[[2]] %>% mutate(var = "cholesterol"), 
  data_summaries[[3]] %>% mutate(var = "protein"), 
  data_summaries[[4]] %>% mutate(var = "sugar")
)

data_summaries
```


4.	Use the function ds_summary_stats() from descriptr to compute the descriptive statistics of this new data frame. Compare the results with the output of your user written function. 

```{r}
descriptr::ds_summary_stats(fastfood %>% select(calories, cholesterol, protein, sugar))
```

5.	Create a box plot for the distribution of calories on the fast food items across restaturants. On the x-axis show the distribution of “calories” and on the y-axis represent the restaurant. Use appropriate labels, coloring, and other graphics best practices.

```{r}
boxplot_calories <- fastfood %>% 
  ggplot(mapping = aes(x = calories, y = restaurant, 
                       fill = restaurant)) + 
  geom_boxplot() + 
  labs(x = "Calories", 
       title = "Distribution of Calories on Fast Food Items by Restaurant") + 
  theme_classic()

boxplot_calories
```


6.	Create a scatter plot that shows the correlation between sugar (x-axis) and protein (y-axis) by restaurants on the same panel. Add a specific color for the observations of each restaurant. Use appropriate labels, coloring, and other graphics best practices.  Save this graph in your environment. Hint: where in the aesthetic mapping should you specify that the graph will vary across restaurants? 

```{r}
scatter_sugar_protein <- fastfood %>% 
  ggplot(mapping = aes(x = sugar, y = protein, 
                       color = restaurant)) + 
  geom_point(size = 1, alpha = 0.5) + 
  labs(x = "Sugar", y = "Protein", 
       title = "Correlation of Sugar and Protein on Fast Food by Restaurant")+ 
  theme_classic()

scatter_sugar_protein
```


7.	Add two lines to your previous scatterplot. A vertical line with the mean of variable sugar, and a horizontal line with the mean of variable cholesterol. Use geom_vline and geom_hline. For these functions you need to specify the xintercept and yintercept as parameters, respectively.

```{r}
mean_sugar <- fastfood$sugar %>% mean(na.rm = TRUE)
mean_protein <- fastfood$protein %>% mean(na.rm = TRUE)

scatter_sugar_protein_improved <- scatter_sugar_protein + 
  geom_vline(xintercept = mean_sugar, color = "navy", linetype = "dashed") + 
  geom_hline(yintercept = mean_protein, color = "maroon", linetype = "dashed") 

scatter_sugar_protein_improved
```


8.	Export this graph using ggsave into your folder. 


```{r}
ggsave(filename = "scatter_protein_sugar.jpg", 
       plot = scatter_sugar_protein_improved)
```

